\documentclass[11pt]{article}
\usepackage{fullpage,amsthm,amsfonts,amssymb,epsfig,amsmath,times,amsthm}
\usepackage[margin=4cm]{geometry}
\usepackage{enumitem}
\usepackage{titlesec}

\setcounter{secnumdepth}{4}

%-------------------------------------
% CS61 students begin copying here

\def\ojoin{\setbox0=\hbox{$\bowtie$}%
  \rule[-.02ex]{.25em}{.4pt}\llap{\rule[\ht0]{.25em}{.4pt}}}
\def\leftouterjoin{\mathbin{\ojoin\mkern-5.8mu\bowtie}}
\def\rightouterjoin{\mathbin{\bowtie\mkern-5.8mu\ojoin}}
\def\fullouterjoin{\mathbin{\ojoin\mkern-5.8mu\bowtie\mkern-5.8mu\ojoin}}

% Relational algebra symbols from ftp://reports.stanford.edu/www/dbgroup_only/latex-macros.html
\newcommand{\select}{\mbox{\Large$\sigma$}}
\newcommand{\cross}{\mbox{$\times$}}
\newcommand{\intersection}{\mbox{$\cap$}}
\newcommand{\intersect}{\mbox{$\cap$}}
\newcommand{\union}{\mbox{$\cup$}}
\newcommand{\join}{\mbox{$\Join$}}
\newcommand{\leftsemijoin}{\mbox{$\mathrel{\raise1pt\hbox{\vrule height5pt
depth0pt width0.6pt\hskip-1.5pt$>$\hskip -2.5pt$<$}}$}}
\newcommand{\rightsemijoin}{\mbox{$\mathrel{\raise1pt\hbox{\hskip-1.5pt$>$\hskip -2.5pt$<$\hskip -1.1pt\vrule height5pt
depth0pt width0.6pt}}$}}
\newcommand{\project}{\mbox{\Large$\pi$}}
\newcommand{\Project}{\mbox{$\Pi$}}
\newcommand{\aggregatefn}{\mbox{\Large$G$}}

% CS61 students end copying here
%-------------------------------------
\newtheorem{theorem}{Theorem}
\newtheorem{claim}[theorem]{Claim}



\begin{document}

\title{Query Languages with Recursion}
\author{Tien Ho}
\maketitle

\begin{center}
%{\bf Liang Wang (lwang107@ucsc.edu)}
\end{center}

\section{Relational Algebra special characters}
\begin{table}[!th]
\begin{tabular}{|l|c|r|}
\hline
selection &  $\select_{cname<cname2 \wedge enr>10000} E$ \\
projection &  $\project_{cname} E$ \\
aggregate function &  $_{g_1,g_2,...} \aggregatefn_{h_1,h_2,...,h_m}$ \\
\hline
\end{tabular}
\caption{Unary operators}
\end{table}

\begin{table}[!th]
\begin{tabular}{|l|c|r|}
\hline
union & 			\union \\
intersection & 		\intersection \\
difference & 		- \\
cartesian product & \cross \\
division & 			$\div$ \\
rename & 			$\rho$ \\
natural join & 		$\bowtie$ \\
theta join & 		$\bowtie_{\theta}$ \\
left semijoin & 	$\leftsemijoin$ \\
right semijoin & 	$\rightsemijoin$ \\ 
left outer join & 	$\leftouterjoin$ \\
right outer join & 	$\rightouterjoin$ \\
full outer join & 	$\fullouterjoin$ \\
antijoin & 			? \\
\hline
\end{tabular}
\caption{Binaryoperators}
\end{table}

\begin{table}[!th]
\begin{tabular}{|l|c|r|}
\hline
Logical AND & 	$\wedge$ \\
Logical OR & 	$\vee$ \\
Logical NOT & 	$\neg$ \\
null & 			$\omega$ \\
\hline
\end{tabular}
\caption{Logic symbols and others}
\end{table}

%Example (Your \LaTeX  skills are likely superior to mine!):
\begin{displaymath}
Grades \leftarrow \project_{(students.ssn, students.name, grades.grade)} \\
(\select_{students.ccn = grades.ccn \wedge grades.assignment = 1} \\
(students \cross grades)) \\
\end{displaymath}

\begin{eqnarray*}
Grades & \leftarrow & \project_{(students.ssn, students.name, grades.grade)}  \\
       &            &(\select_{(students.ssn, students.name, grades.grade)}  \\
       &            &          (students \cross grades)) 
\end{eqnarray*}

\section{Introduction}

\section{Negation in Logic Programming}

\section{$fixpoint$ Queries}
author: Liang\\
introduction, belong to LFP
\subsection{Semantics}
describe
\subsection{Expressiveness}
discuss
\subsection{Complexity}
P,
important theorems and cited proofs
\subsection{Problems}
discuss

\section{CALC-$\mu$}
author: Liang\\
introduction, belong to PFP
\subsection{Semantics}
describe
\subsection{Expressiveness}
discuss
\subsection{Complexity}
PSPACE-complete,
important theorems and cited proofs
\subsection{Problems}
discuss

\section{The $while$ Language}
\subsection{Semantics}
\subsection{Expressiveness}
\subsection{Complexity}
PSPACE
\subsection{Problems}

\input{StableModel}

\begin{thebibliography}{9}
  \bibitem{lifschitz0} 
  Vladimir Lifschitz. 
  What is answer set programming? 
  In \textit{Proceedings of the AAAI Conference on Artificial Intelligence}, pages 1594-1597. 
  MIT Press, 1988.

  \bibitem{lifschitz1} 
  Vladimir Lifschitz. 
  Datalog Programs and their stable models.
  In \textit{Proceedings of the AAAI Conference on Artificial Intelligence}, pages 1594-1597. 
  MIT Press, 1988.

  \bibitem{gelfond} 
  Michael Gelfond and Vladimir Lifschitz. 
  The stable model semantics for logic programming.  
  In \textit{Proceedings of International Logic Programming Conference and Symposium}, pages 1070-1080. 
  MIT Press, 2008.

  \bibitem{ferraris} 
  Paolo Ferraris and Vladimir Lifschitz. 
  Mathematical foundations of answer set programming.  
  In \textit{We Will Show Them! Essays in Honour of Dov Gabbay}, pages 615-664. 
  King's College Publications, 2005.

  \bibitem{Marek} 
  Victor Marek and Miroslaw Truszczynski. 
  Stable models and an alternative logic programming paradigm.  
  In \textit{The Logic Programming Paradigm: a 25-Year Perspective}, pages 375-398. 
  Springer Verlag, 1999.

  \bibitem{lee} 
  Paolo Ferraris, Joohyung Lee, and Vladimir Lifschitz. 
  A new perspective on stable models.  
  In \textit{Proceedings of International Joint Conference on Artificial Intelligence (IJCAI)}, pages 372-379. 
  2007.

  \bibitem{lifschitz2} 
  Vladimir Lifschitz. 
  Thirteen definitions of a stable model.  
  In \textit{Lecture Notes in Computer Science}, vol 6300. 
  Springer Verlag, 2010.
\end{thebibliography}

\end{document}